---
title: "Lab 13: Creating a detailed report with explanations from the NEON MAG data"
author: "Marina Duclos"
format:
  html:
    toc: true
    toc_float: true
    code-fold: true
    theme: modern
    embed-resources: true
execute: 
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(vegan)

data <- read_csv("NEON_soilMAGs_soilChem (1).csv")
```

# Canonical Correspondance Analysis (CCA) & Variance Paritioning

```{r}
# Load libraries
library(tidyverse)
library(vegan)
library(pheatmap)

# Load data
data <- read_csv("NEON_soilMAGs_soilChem (1).csv")

# Create phylum-level abundance matrix
taxa_matrix <- data %>%
  filter(!is.na(phylum)) %>%
  count(genomicsSampleID, phylum) %>%
  pivot_wider(names_from = phylum, values_from = n, values_fill = 0)

# Select environmental variables
env_data <- data %>%
  select(genomicsSampleID, soilMoisture, soilTemp, soilInWaterpH, soilInCaClpH, elevation) %>%
  distinct()

# Join and format
merged <- inner_join(env_data, taxa_matrix, by = "genomicsSampleID")
rownames(merged) <- merged$genomicsSampleID

# Split into environmental and taxonomic data
env <- merged %>% select(soilMoisture, soilTemp, soilInWaterpH, soilInCaClpH, elevation)
taxa <- merged %>% select(-genomicsSampleID, -soilMoisture, -soilTemp, -soilInWaterpH, -soilInCaClpH, -elevation)

# Ensure numeric and complete data
taxa <- taxa %>% mutate(across(everything(), as.numeric))
complete_rows <- complete.cases(env) & complete.cases(taxa)
env <- env[complete_rows, ]
taxa <- taxa[complete_rows, ]

# Match rownames
common_ids <- intersect(rownames(env), rownames(taxa))
env <- env[common_ids, ]
taxa <- taxa[common_ids, ]

# Run CCA
cca_result <- cca(taxa ~ soilMoisture + soilTemp + soilInWaterpH + soilInCaClpH + elevation, data = env)

# Plot CCA biplot
plot(cca_result, main = "CCA: Taxonomic Composition vs. Soil Chemistry")

# Add environmental vectors
envfit_result <- envfit(cca_result, env, permutations = 999)
plot(cca_result)
plot(envfit_result, p.max = 0.05, col = "blue")

# Test significance
anova(cca_result)              # Overall model
anova(cca_result, by = "term") # Individual variables

# Variance partitioning: split into two explanatory tables
chemistry <- env %>% select(soilMoisture, soilTemp)
pH_elevation <- env %>% select(soilInWaterpH, soilInCaClpH, elevation)

varpart_result <- varpart(taxa, chemistry, pH_elevation)
plot(varpart_result, cutoff = 0.01,
     bg = c("lightblue", "lightgreen", "lightpink", "lightgray"))

# Correlation heatmap of taxa vs. environment
taxa_scaled <- decostand(taxa, method = "hellinger")
cor_matrix <- cor(taxa_scaled, env)
pheatmap(cor_matrix, cluster_rows = TRUE, cluster_cols = TRUE, display_numbers = TRUE)
```


## Exercise 3

### Alpha Diversity Metrics 

```{r}
# Compute alpha diversity metrics
diversity_data <- taxa %>%
  mutate(shannon = diversity(., index = "shannon"),
         simpson = diversity(., index = "simpson"),
         richness = specnumber(.))

# Combine with environmental data
div_summary <- bind_cols(env[rownames(diversity_data), ], diversity_data)
```

### Beta Diversity (NMDS)

```{r}
# Run NMDS
nmds_result <- metaMDS(taxa, distance = "bray", k = 2, trymax = 100)
plot(nmds_result, main = "NMDS: Beta Diversity")
```

### Environmental Gradient Overlay

```{r}
# Fit environmental vectors to NMDS
nmds_envfit <- envfit(nmds_result, env, permutations = 999)
plot(nmds_result)
plot(nmds_envfit, p.max = 0.05, col = "darkgreen")
```



## Exercise 4

### Shanon Diversity

Measures both richness and evenness of taxa in a sample. Higher values indicate more diverse communities


### Simpson Diversity 

Emphasizes dominance; lower values suggest one or few taxa dominate the community


### Richness

Counts the number of unique taxa present in a sample — a simple measure of biodiversity

### Canonical Correspondence Analysis (CCA)

An ordination method that relates community composition to environmental gradients using constrained axes


### Non-metric Multidimensional Scaling (NMDS)

An unconstrained ordination method that visualizes community similarity based on dissimilarity matrices like Bray-Curtis

### Envit 

Projects environmental variables as vectors onto ordination plots, showing direction and strength of influence

### Variance Partitioning (varpart)

Partitions variation in community composition across multiple environmental subsets to assess their relative contributions









